\input{./mac/head.tex}
\input{./mac/maths.tex}

\Bdc{Notes on Probability Theory}

\section{Measures}

Suppose that \((A_n)_{n \in \setnat}\) is a countably infinite sequence of sets. The sequence is \emph{increasing} if
\(A_n \subseteq A_{n + 1}\) for each \(n \in \setnat\), and \emph{decreasing} if \(A_{n + 1} \subseteq A_n\) for each
\(n \in \setnat\). The sequence is \emph{monotone} if it is increasing or decreasing.

If \((A_n)_{n \in \setnat}\) is increasing, the \emph{increasing union thereof} is
\[
  \bigcup_{n \in \setnat} A_n;
\]
if \((A_n)_{n \in \setnat}\) is decreasing, the \emph{decreasing intersection thereof} is
\[
  \bigcap_{n \in \setnat} A_n.
\]

The \emph{limit inferior of the countably infinite sequence \((A_n)_{n \in \setnat}\)} is
\[
  \liminf_{n \to \infty} A_n = \bigcup_{n = 0}^\infty \bigcap_{i = n}^\infty A_i,
\]
and the \emph{limit superior thereof} is
\[
  \limsup_{n \to \infty} A_n = \bigcap_{n = 0}^\infty \bigcup_{i = n}^\infty A_i.
\]

\subsection{Topological Spaces and Metric Spaces}

\Bdf
  Let \(X\) be a set. Then \(\mathcal{T} \subseteq \mathcal{P}(X)\) is a \emph{topology on \(X\)} if and only if
  \begin{enumerate}
    \item \(\nset \in \mathcal{T}\);
    \item \(X \in \mathcal{T}\);
    \item \(\mathcal{T}\) is closed under finite intersections; and
    \item \(\mathcal{T}\) is closed under unions.
  \end{enumerate}

  A \emph{topological space} is an ordered pair \((X, \mathcal{T})\) wherein \(X\) is a set and \(\mathcal{T}\) is a
  topology thereon, and a subset of \(X\) is \emph{open} if it is in \(\mathcal{T}\).
\Edf

A subset of \(X\) is \emph{closed} if and only if the relative complement thereof is open. For each \(A \subseteq X\),
the \emph{closure of \(A\)} is the intersection of all closed sets which include \(A\). A subset of \(X\) is
\emph{dense} if and only if the closure thereof is \(X\). The topological space is \emph{separable} if and only if there
exists a countable dense subset of \(X\).

For each \(\mathcal{X} \subseteq \mathcal{P}(X)\), there exists a smallest topology on \(X\) which includes
\(\mathcal{X}\), referred by as the \emph{topology generated by \(\mathcal{X}\)}.

Let \((X_i, \mathcal{T}_i)_{i \in I}\) be topological spaces with \(I\) as an index set. Let \(\mathcal{X}\) be the
cartesian product of \((X_i)_{i \in I}\); i.e.,
\[
  \mathcal{X} = \prod_{i \in I} X_i.
\]
For each \(i \in I\), let \(\map{\text{pr}_i}{\mathcal{X}}{X_i}\) denote the \(i\)-th projection on \(\mathcal{X}\);
i.e.,
\[
  \text{pr}_i(x) = x_i
\]
wherein \(x = (x_i)_{i \in I}\). The \emph{product topology on \(\mathcal{X}\)} is the topology generated by
\[
  \set{\text{pr}_i^{-1}[U] : U \in \mathcal{T}_i \land i \in I}.
\]

\Bdf
  Let \(X\) be a set. Then \(\map{d}{X^2}{\setreal}\) is a \emph{metric on \(X\)} if and only if
  \begin{enumerate}
    \item \(d(x, x) = 0\) for each \(x \in X\);
    \item \(d(x, y) > 0\) if \(x \ne y\) for each \(x, y \in X\);
    \item \(d(x, y) = d(y, x)\) for each \(x, y \in X\); and
    \item (\emph{triangle inequality}) \(d(x, y) + d(y, z) \ge d(x, z)\) for each \(x, y, z \in X\).
  \end{enumerate}

  A \emph{metric space} is an ordered pair \((X, d)\) wherein \(X\) is a set and \(d\) is a metric thereon.
\Edf

Given \(\varepsilon \in \posreal\) and \(x \in X\), the \emph{open \(\varepsilon\)-ball of \(x\) in \(X\)} is
\[
  B_d(x, \varepsilon) = \set{y : d(x, y) < \varepsilon}.
\]
The \emph{metric topology on \(X\) induced by \(d\)} is
\[
  \set*{U \subseteq X : \forall \, x \in U \, \big(\exists \, \varepsilon \in \posreal \, \left(B_d(x, \varepsilon)
  \subseteq U\right)\big)}.
\]

A topological space \((X, \mathcal{T})\) is \emph{metrisable} if and only if there exists a metric on \(X\) whereby
\(\mathcal{T}\) is the induced metric topology.

\subsection{\texorpdfstring{\(\sigma\)}{Sigma}-Algebras}

\Bdf
  Let \(X\) be a set. Then \(\varSigma \subseteq \mathcal{P}(X)\) is a \emph{\(\sigma\)-algebra on \(X\)} if and only if
  \begin{enumerate}
    \item \(\varSigma \ne \nset\);
    \item \(\varSigma\) is closed under complementation; and
    \item \(\varSigma\) is closed under countable unions.
  \end{enumerate}

  A \emph{measurable space} is an ordered pair \((X, \varSigma)\) wherein \(X\) is a set and \(\varSigma\) is a
  \(\sigma\)-algebra thereon, and a subset of \(X\) is \emph{measurable} if it is in \(\varSigma\).
\Edf

It follows that \(X \in \varSigma\) and \(\nset \in \varSigma\) for each \(\sigma\)-algebra on \(X\) and that each
\(\sigma\)-algebra is closed under countable intersections. Hence, the smallest \(\sigma\)-algebra on \(X\) is \(\set{X,
\nset}\) and the largest thereon is \(\mathcal{P}(X)\).

A \(\sigma\)-algebra is closed under both increasing unions and decreasing intersections.

For each \(P \in \mathcal{P}(X)\), there exists a smallest \(\sigma\)-algebra on \(X\) which includes \(P\), referred to
as the \emph{\(\sigma\)-algebra generated by \(P\)} and denoted by \(\sigma(P)\).

A \emph{\(\pi\)-system on \(X\)} is a set of subsets of \(X\) which is closed under finite intersections. A
\emph{Dynkin system on \(X\)} is a set of subsets of \(X\) which contains \(X\) and is closed under both proper
differences and increasing unions.

\Bth[Sierpiński--Dynkin's \(\pi\)--\(\lambda\) theorem]
  Let \(\mathcal{C}\) be a \(\pi\)-system on \(X\), and let \(\mathcal{D}\) be a Dynkin system thereon.

  If \(\mathcal{C} \subseteq \mathcal{D}\), then \(\sigma(\mathcal{C}) \subseteq \mathcal{D}\).
\Eth
\Bpr
  First, \(\mathcal{C}\) is a \(\sigma\)-algebra if and only if it is also a Dynkin system.

  Assume that \(\mathcal{D}\) is the smallest Dynkin system which includes \(\mathcal{C}\). It then suffices to show
  that \(\mathcal{D}\) is also a \(\pi\)-system. Hence, we need to show that \(A \cap B \in \mathcal{D}\) for any
  \(A, B \in \mathcal{D}\).

  To demonstrate the above proposition, first define \(\mathcal{X}_B = \set{A \subseteq X : A \cap B \in \mathcal{D}}\)
  for each \(B \in \mathcal{C}\). Then each \(\mathcal{X}_B\) is a Dynkin system including \(\mathcal{C}\), and so it
  includes \(\mathcal{D}\). Thus, \(A \cap B \in \mathcal{D}\) for each \(A \in \mathcal{D}\) and \(B \in \mathcal{C}\).
  Next, define \(\mathcal{X}'_A = \set{B \subseteq X : A \cap B \in \mathcal{D}}\) for each \(A \in \mathcal{D}\).
  Similarly, we note that each \(\mathcal{X}'_A\) includes \(\mathcal{D}\).

  Therefore, the theorem holds.
\Epr

\Bdf
  Let \((X, \mathcal{T})\) be a topological space. The \emph{Borel \(\sigma\)-algebra of \((X, \mathcal{T})\)}, denoted
  by \(\mathcal{B}(X, \mathcal{T})\), is the \(\sigma\)-algebra generated by \(\mathcal{T}\).

  A \emph{Borel set of \((X, \mathcal{T})\)} is an element of \(\mathcal{B}(X, \mathcal{T})\).
\Edf

Let \((X_i, \varSigma_i)_{i \in I}\) be measurable spaces with \(I\) as an index set. The \emph{product
\(\sigma\)-algebra of \((\varSigma_i)_{i \in I}\)} is
\[
  \bigotimes_{i \in I} \varSigma_i = \sigma\left(\set*{A_i \times \prod_{j \in I \land j \ne i} X_j :
  A_i \in \varSigma_i \land i \in I}\right).
\]

\Blm
  Let \((X_i)_{i \in I}\) be separable metric spaces with \(I\) as an index set. Then
  \[
    \mathcal{B}\left(\prod_{i \in I} X_i\right) = \bigotimes_{i \in I} \mathcal{B}(X_i).
  \]
\Elm

\Edc
