\input{./mac/head.tex}
\input{./mac/maths.tex}

\Bdc{Notes on Syntax}

\section{Syntactic Categories}

The \emph{syntactic category} of a word, also known as its \emph{part of speech} or \emph{word class}, is determined by
its \emph{distribution}. We note syntactic categories including
\begin{enumerate}
  \item \emph{nouns}, denoted by N;
  \item \emph{verbs}, denoted by V;
  \item \emph{adjectives}, denoted by Adj;
  \item \emph{adverbs}, denoted by Adv;
  \item \emph{prepositions}, denoted by P;
  \item \emph{determiners}, denoted by D;
  \item \emph{conjunctions}, denoted by Conj;
  \item \emph{complementisers}, denoted by C;
  \item \emph{tense markers}, denoted by T; and
  \item \emph{negation}, denoted by Neg.
\end{enumerate}

\subsection{Distribution}

\subsubsection{Morphological Distribution}

The smallest meaningful units in language are referred to as \emph{morphemes}. A \emph{free morpheme} can occur
independently, and a \emph{bound morpheme} or \emph{affix} must attach to another unit, referred to as the \emph{stem}.
An affix attached at the end of the stem is a \emph{suffix}; at the start, a \emph{prefix}; inside, an \emph{infix}; and
at both the start and the end, a \emph{circumfix}.

The stem required by an affix is referred to as its \emph{complement}. An affix requires its complement to be of a
certain syntactic category, a process referred to as \emph{c-selection} or \emph{category selection}. That affixes
c-selects complements of certain syntactic categories form the \emph{morphological distribution} of those categories.

A \emph{lexeme} is a unit of meaning underlying a set of word forms related by inflection. An affix which modifies the
lexeme of its complement is a \emph{derivational morpheme}, and one which retains the lexeme thereof is an
\emph{inflectional morpheme}.

\Bxm
  The English suffix \textit{-ise} is a derivational morpheme and c-selects N or Adj.
\Exm

\subsubsection{Syntactic Distribution}

A syntactic category occurs in certain syntactic structures, referred to as its \emph{syntactic distribution}.

\subsubsection{Complementary Distribution}

Two words are \emph{in complementary distribution} if their distributions do not intersect. Complementary distribution
implies identical syntactic category.

\subsection{Classification of Syntactic Categories}

We classify the syntactic categories noted above in two ways.

\subsubsection{Open and Closed Classes}

Syntactic categories that permit new members are \emph{open classes}, and those that do not are \emph{closed classes}.
N, V, Adj, and Adv are open classes; and P, D, Conj, C, T, and Neg are closed classes.

\subsubsection{Lexical and Functional Categories}

Syntactic categories with full semantic content are \emph{lexical categories}, and those without are \emph{functional
categories}. Lexical categories coincide with open classes, and functional categories coincide with closed classes.

\subsection{Subcategories and Features}

D and T have \emph{subcategories}. The subcategories of D are
\begin{enumerate}
  \item \emph{articles},
  \item \emph{deictic articles},
  \item \emph{quantifiers},
  \item \emph{numerals},
  \item \emph{possessive pronouns}, and
  \item some \textit{wh-}question words;
\end{enumerate}
and the subcategories of T are
\begin{enumerate}
  \item \emph{auxiliaries},
  \item \emph{modals}, and
  \item non-finite tense marker.
\end{enumerate}

\Edc
